---
title: "Diversity Index for Maryland Cities"
date: "Nov 8, 2025"
author: Lauren Frank
---
Lauren Frank

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
```

## Overview

In this exercise, we will calculate the **USA TODAY Diversity Index** for Maryland cities with a population of at least 10,000 using 2020 Census race and ethnicity data. The Diversity Index reflects the probability that two randomly selected individuals from a city are of different races or ethnicities.

A score of 0 indicates no diversity (everyone is the same race/ethnicity), while a score closer to 100 indicates higher diversity. The **most diverse city** has the highest index, reflecting a balanced mix of several racial and ethnic groups. The **least diverse** city likely has one overwhelmingly dominant racial/ethnic group.

## Task 1: Load and Describe the Data (3 points)

Run the following code to load the data, then examine it and describe what it contains. Based on the raw numbers alone, which cities stand out as diverse to you?

```{r load-data}
md_cities <- read_csv("https://raw.githubusercontent.com/dwillis/jour405/refs/heads/main/data/usat_diversity_index.csv")

# Preview structure
View(md_cities)
```
Based on the data, Baltimore City stands out as a diverse city. Among the total population of 622,454, over half of the population is Black (391,122) and 14,123 of the population identifies with two or more races. Another city that stands out is Germantown. Among the total population of 89,826, about just under half of the population is white (41,664), and about 1/4 of the population is Black (20,532). 
)

## Task 2: Calculate Race Percentages

```{r race-percentages}
md_cities <- md_cities |>
  mutate(
    White_pct = White / TotalPop,
    Black_pct = Black / TotalPop,
    Native_pct = Native / TotalPop,
    Asian_pct = Asian / TotalPop,
    NHPI_pct = NHPI / TotalPop
  )
```

## Task 3: Square the Race Percentages

We need to square the race percentages in order to calculate the probability that any two people chosen at random from a given census area are of different races or are Hispanic or not.

```{r square-race}
md_cities <- md_cities |>
  mutate(
    White_pct_sq = White_pct^2,
    Black_pct_sq = Black_pct^2,
    Native_pct_sq = Native_pct^2,
    Asian_pct_sq = Asian_pct^2,
    NHPI_pct_sq = NHPI_pct^2
  )
```


## Task 4: Calculate Race Probabilities (3 points)

The race probability is the sum the squared probabilities for each of the races. It represents the probability that two people randomly chosen from the same city will be of the same race. Which city has the highest race probability, and why do you think that is

```{r race-prob}
md_cities <- md_cities |>
  mutate(
    Race_Prob = White_pct_sq + Black_pct_sq + Native_pct_sq + Asian_pct_sq + NHPI_pct_sq
  ) |> 
  arrange(desc(Race_Prob)) |> 
  select(Name, Race_Prob, everything())

View(md_cities)
```

Summerfield has the highest probability (0.87152625) that two people randomly chosen from the same city will be of the same race. I think this is the case because of the population (12,403), 11,565 of those residents are Black. This means there is a high probability that the two people randomly chosen will both be Black. 

## Task 5: Calculate Ethnicity Percentages and Their Squares (3 points)

Hispanic origin is not a race but an ethnicity or nationality – and a separate Census question. The probability someone is Hispanic or not Hispanic must be calculated separately from race, and Hisp_Prob is the probability that two randomly selected people will be Hispanic or not. That's an important distinction, because unlike race probability, this measures the likelihood of a binary question: that both random people will be Hispanic. What do you notice about the cities with the *lowest* Hisp_Prob scores?

```{r ethnicity}
md_cities <- md_cities |>
  mutate(
    HispTotal = HISPANIC + NotHispanic,
    Hisp_pct = HISPANIC / HispTotal,
    NotHisp_pct = NotHispanic / HispTotal,
    Hisp_pct_sq = Hisp_pct^2,
    NotHisp_pct_sq = NotHisp_pct^2,
    Hisp_Prob = Hisp_pct_sq + NotHisp_pct_sq
  ) |> 
  arrange(Hisp_Prob)

View(md_cities)
```

I noticed that the cities with the lowest Hisp_Prob scores have lower Race_Prob scores. For example, Chillum has a Hisp_Prob score of 0.5063968 and a Race_Prob score of 0.25411432. However, Kettering has a Hisp_Prob score of 0.9797471 and a Race_Prob score of 0.82552476. 

## Task 6: Multiply Race and Ethnicity Probabilities

Here we multiply the two probabilities together to get the probability that any two random people have the SAME race and ethnicity.

```{r combined-prob}
md_cities <- md_cities |>
  mutate(
    RaceEth_Prob = Race_Prob * Hisp_Prob
  )
```


## Task 7: Calculate Diversity Index (3 points)

The final step is to calculate the probability that two people randomly chosen WILL NOT BE OF THE SAME RACE OR ETHNICITY. In other words, the highest index scores will be cities where those two random people will be from different backgrounds. Put simply, more diverse. Which cities are the top 3, and where are they located? Look at the bottom of the list - which city is the least diverse, and where is it?


```{r diversity-index}
md_cities <- md_cities |>
  mutate(
    Diversity_Index = (1 - RaceEth_Prob) * 100
  ) |> 
  arrange(desc(Diversity_Index))
  
View(md_cities)
```

Least Diverse: Lake Shore, Lake Arbor and Largo.
Most Diverse: Langley Park, Wheaton and Adelphi.

### Pitch a story (8 points)

Given the data that you've produced here, write a headline and lede of the most interesting story from this exercise, and add below that details on how you would report it out: who you'd talk to and the questions you'd ask. When you describe the Diversity Index values, do so in a similar way to how USA Today did; the raw values themselves won't mean much to readers.

Headline: Most Diverse Cities in PG and Montgomery County 
Lede: Langley Park, Wheaton and Adelphi are among the top scoring cities in the diversity index, according to recent data. 

Details on how you would report it out/ who you'd talk to and the questions you'd ask:
- I would report on the Maryland cities with a population of at least 10,000 using 2020 Census race and ethnicity data. 
- I would include the data from the NotHisp_pct. I noticed that the the lower the NotHisp_pct is, the higher the Diversity Index score is. For example, Langley Park had a NotHisp_pct score of 0.1893677 and a Diversity Index Score of 97.18020. Whereas, Kettering had a NotHisp_pct score of 0.989768 and a Diversity Index Score of 19.11945.
- It would be important to talk to someone who has knowledge about diversity, urbanization or politcs? I looked at a similar article from U.S.news and they had a source who was a "leader demographer and senior fellow at the Brookings Institution." The article also interviews an urban sociologist at the University of Washington. So in the case of my article, I could talk to professionals in the cities that scored the highest and/or lowest in Diversity. Maybe even a professor at UMD or other nearby colleges and institutions. 

