---
title: "HW7_MandMs"
date: 10/1/2025
---

In this assignment, you'll be picking up where you left off with the M&Ms activity. You'll load the data and perform some basic statistical analysis. Follow the steps below to complete the assignment.

### Task 1: Load the tidyverse

```{r}
library(tidyverse)
```

### Task 2: Load an M&Ms dataset from https://raw.githubusercontent.com/dwillis/jour405/refs/heads/main/data/mandms.csv
```{r}
MM <- read_csv ("https://raw.githubusercontent.com/dwillis/jour405/refs/heads/main/data/mandms.csv")
```

### Task 3: Sample 5 rows of data from the dataset, and then sample 10 rows, saving each to separate dataframes called sample5 and sample10. Be sure to use set.seed(1234) before you start.
```{r}
set.seed(1234)
sample5 <- MM |> 
  sample_n(5)
sample10 <- MM |>
  sample_n(10)
```


### Task 4: Calculate the mean, median and standard deviation of red M&Ms from each of the samples.
```{r}
sample5 |> summarize(mean= mean(red), median = median(red), sd = sd(red))
sample10 |> summarize(mean = mean(red), median = median(red), sd = sd(red))
```


### Task 5: Calculate the mean, median and standard deviation of red M&Ms from the population

```{r}
MM |> summarize(mean = mean(red), median = median(red), sd = sd(red))
```

### Task 6: Make three histograms: one each for sample5, sample10 and the original dataset. Include a headline. Use a consistent binwidth.

```{r}
sample5 |> 
  ggplot() + 
  geom_histogram(aes(x = red), binwidth = 4) +
  labs(title="Distribution of Red M&Ms", x = "Number of Red M&Ms", y = "Number of people with M&Ms") 
sample10 |>
  ggplot() + 
  geom_histogram(aes(x = red), binwidth = 4) +
  labs(title="Distribution of Red M&Ms", x = "Number of Red M&Ms", y = "Number of people with M&Ms")
MM |>
  ggplot() + 
  geom_histogram(aes(x = red), binwidth = 4) +
  labs(title="Distribution of Red M&Ms", x = "Number of Red M&Ms", y = "Number of people with M&Ms")
```


### Task 7: Answer questions

Compare the three sets of means, standard deviations and histograms. Which are more helpful to you in understanding the accuracy of the samples?
A: Among the three sets of means, sd's and histograms, the population values are most helpful in understanding the accuracy of the samples because it provides a more spread out or distributed version of the data that covers a wide range of values. A population histogram shows the entire population's distribution rather than a small portion of it, which may not accurately represent.

What does this exercise tell you about the relationship between sample size and estimate accuracy?
A: This exercise tells me that the larger sample sizes produce more accurate estimates that are closer to the actual population. The more data available, the less likely the outliers/extreme values will impact the data and accuracy. 

How does random sampling introduce variation into statistical estimates?
A: Random sampling introduces variation into statistical estimates because each sample chosen will be different. 

If you repeated this process, would the samples always be identical? Why or why not?
A: The samples would not always be identical because the randomness is what creates differences in sample statistics compared to one another and the population.

When finished, save your work, switch to GitHub Desktop, then add, commit and push your changes to GitHub and submit the URL of the notebook in ELMS.
