---
title: "Car Thefts Significance"
author: "Derek Willis/Rob Wells"
date: 10-27-2025
---

name: Lauren Frank


Your editor wants to know if there have been recent changes in the number of car thefts in Prince George's County. Using monthly figures and significance testing, you'll come up with an answer and decide whether there's a story here. As with the previous homework, provide answers in this notebook when prompted and you will have to fill in values where you see REPLACE_ME. Get started by loading the tidyverse in a new R block.

### Task 1: Load the tidyverse

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
```


### Task 2: Load the Prince George's county crime data and generate car thefts totals by month

```{r}
# Read the data
car_thefts <- read_csv ("https://raw.githubusercontent.com/dwillis/jour405/refs/heads/main/data/car_thefts_month.csv") |>
  
arrange(month) # Sort by date

# Basic inspection of the data
glimpse(car_thefts)
```

### Task 3: Let's make a chart and describe the results.

We'll make a line chart using geom_line() to try and illustrate any trend over time. 

```{r}

car_thefts |> 
ggplot(aes(x = month, y = total)) +
  geom_line(color = "steelblue", size = 1) +
  geom_point(color = "darkblue", size = 3) +
  labs(
    title = "Monthly Car Thefts (July 2023 - February 2025)",
    subtitle = "Is there a trend over time?",
    x = "Month",
    y = "Number of Car Thefts"
  ) +
  theme_minimal() +
  theme(
    plot.title = element_text(face = "bold", size = 14),
    axis.title = element_text(face = "bold")
  )

```


**Q3:** What's your initial reaction to the change in car thefts? Is there a trend?

I'm surpised that some of the lowest number of car thefts occured in the April. I would think that there would be less car thefts in the winter months like Jan. or Feb. I noticed that in both 2023 and 2024, it appears there was a spike in car thefts in the fall months around Sept/Oct. I also noticed that there was a significant drop of car thefts from Dec. to Jan. 

Generally, the trend I noticed is that number of car thefts were higher during the Summer (around July) and in the fall, with noticably higher peaks in Dec. 2023 and Oct 2024. 

### Task 4: Add the mean, standard error and calculate a margin of error based on a 95% confidence interval

Let's add some stats to that process - mean and standard deviation - and calculate the standard error and 95% confidence intervals. We only have a small number of months, so we'll calculate the margin of error to account for that. 

**INSERT CORRECT VARIABLE AT "REPLACE ME"**

```{r}

theft_summary <- car_thefts |> 
  summarize(
    mean_thefts = mean(total),
    sd_thefts = sd(total),
    months = n(),
    se_thefts = sd_thefts / sqrt(months),
    margin_error = qt(0.975, df = months - 1) * se_thefts,  # Using t-distribution for small sample
    ci_lower = mean_thefts - margin_error,
    ci_upper = mean_thefts + margin_error
  )

# Display the confidence interval
theft_summary
```




### Task 5: Visualize the data

Let's make a better chart showing the changes and the statistics we produced in Task 4. Does this change anything about your answer from Task 3?

```{r}
# Make a better chart
car_thefts |> 
ggplot(aes(x = month, y = total)) +
  geom_line(color = "steelblue", alpha = 0.7) +
  geom_point(size = 2) +
  geom_hline(yintercept = theft_summary$mean_thefts, linetype = "dashed", color = "red", size = 1) +
  geom_ribbon(aes(ymin = theft_summary$ci_lower, ymax = theft_summary$ci_upper), 
              fill = "red", alpha = 0.2) +
  annotate("text", x = min(car_thefts$month), 
           y = theft_summary$mean_thefts + 10, 
           label = paste0("Mean: ", round(theft_summary$mean_thefts, 1)), 
           hjust = 0, color = "red") +
  labs(
    title = "Monthly Car Thefts with 95% Confidence Interval for the Mean",
    x = "Month",
    y = "Number of Car Thefts"
  ) +
  theme_minimal()

```

### Task 6: Describe the confidence interval.

- What does the 95% confidence interval show here? The 95% confidence intervals shows that 95% of the confidence intervals created capture the true value, and it gives the range of plausible values for the parameter of car thefts. 

- What figure(s) accurately reflects the number of car thefts in a typical month? The mean shows the “typical” number of car thefts per month. The 95% confidence interval show the plausible range around that mean. 

- Propose a headline about the trend in car thefts that is focused on the present and future, but informed by the data you see above. ** "Car Thefts Return to Typical Levels in Prince George’s County despite seasonal spikes”


### Task 7: Is the recent trend significant?

It _looks_ like car thefts are decreasing lately. But is that change real? We'll split the monthly data into two halves, and then ask if whether there's a statistically significant difference between these periods. To do that, we'll use a t-test with our alpha at 0.05 and develop a null hypothesis and an alternative hypothesis for this scenario.

Null Hypothesis: there is no real difference between the two periods
Alternative Hypothesis: there is a real difference between the two periods

**INSERT CORRECT VARIABLES AT "REPLACE ME"**

```{r}

# Find the halfway point and add a column placing each month in the first or second halves

midpoint <- ceiling(nrow(car_thefts) / 2)
car_thefts <- car_thefts |> 
  mutate(period = if_else(row_number() <= midpoint, "First Half", "Second Half"))

# Perform t-test comparing the two time periods - this is a two-sided test because we have two periods
t_test_periods <- t.test(total ~ period, data = car_thefts)

# Display results
t_test_periods

# Visualize the comparison
period_summary <- car_thefts |> 
  group_by(period) |> 
  summarize(
    mean_thefts = mean(total),
    sd_thefts = sd(total),
    months = n(),
    se_thefts = sd_thefts / sqrt(months),
    margin_error = qt(0.975, df = months - 1) * se_thefts, # look at how we calculated this above
    ci_lower = mean_thefts - margin_error,
    ci_upper = mean_thefts + margin_error
  )

period_summary |> 
ggplot(aes(x = period, y = mean_thefts, fill = period)) +
  geom_col() +
  geom_errorbar(aes(ymin = ci_lower, ymax = ci_upper), width = 0.2) +
  geom_text(aes(label = round(mean_thefts, 1)), vjust = -0.5, size = 4) +
  labs(
    title = "Comparing Car Thefts: First Half vs. Second Half of Data Period",
    subtitle = paste0("p-value = ", round(t_test_periods$p.value, 3)),
    x = "Time Period",
    y = "Average Number of Car Thefts"
  ) +
  scale_fill_manual(values = c("First Half" = "#66c2a5", "Second Half" = "#fc8d62")) +
  theme_minimal() +
  theme(legend.position = "none")

```

### Task 8: Reflection

Given the above chart, is the difference between the two periods statistically significant? Can we reject the null hypothesis? 

Given the chart above, there is no statistically significant difference between the first and second periods. We can't reject the null hypothesis because the p-value (0.18) is greater than the alpha (0.05). The null hypothesis stated that there is no statistical difference between the two periods.

How would you revise your headline from Task 6, if at all? What would you tell your editor, and what would you expect her to ask you? 

- Revised headline: "
- I would tell my editor that although there’s a visual decline in recent months, the difference isn’t statistically significant enough. I might expect her to ask me whether there's a real decrease either shown through the recent or long-term data. 

When you are finished, add, commit and push your notebook and submit the URL in ELMS.
