---
title: "HW6_Spotify_Songs"
---

In this assignment, you will be working with a dataset of songs played on Spotify in 2020, and you'll be writing code to explore the range of danceability scores in the data, using standard deviation and histograms. You'll also write up your findings.

### Task 1: Load the tidyverse

```{r}
library(tidyverse)
```

### Task 2: Load the Spotify Songs dataset from https://raw.githubusercontent.com/dwillis/jour405/refs/heads/main/data/spotify_2020.csv

```{r}
spotify_songs <- read_csv("https://raw.githubusercontent.com/dwillis/jour405/refs/heads/main/data/spotify_2020.csv")
```
### Task 3: Calculate the mean and standard deviation of the danceability scores

Write code to calculate the mean and standard deviation of the danceability scores in the Spotify dataset. Display those.

```{r}
spotify_songs |> summarize(mean = mean(danceability), sd = sd(danceability))
```

### Task 4: Create a histogram of the danceability scores

Write code to create a histogram of the danceability scores in the Spotify dataset, based on the example we did in class.

```{r}
spotify_songs |> 
  ggplot() + 
  geom_histogram(aes(x = danceability), binwidth = 0.05)
```
``
### Task 5: Add a line to the histogram for the mean danceability score

Take the same code and add a vertical line to the histogram for the mean danceability score, based on the example in class.

```{r}
spotify_songs |> 
  ggplot() + 
  geom_histogram(aes(x = danceability), binwidth = 0.05) + 
  geom_vline(aes(xintercept = mean(danceability)), color = "red", linetype = "dashed", size = 1)
```

### Task 6: Add two lines to the histogram to show one standard deviation

Take the previous code and add two lines to the histogram for the 1st standard deviation below and above the mean score.

```{r}
spotify_songs |> 
  ggplot() + 
  geom_histogram(aes(x = danceability), binwidth = 0.05 ) + 
  geom_vline(aes(xintercept = mean(danceability)), color = "red", linetype = "dashed", size = 1) +
  geom_vline(aes(xintercept = mean(danceability) - sd(danceability)), color = "blue", linetype = "dashed", size = 1) +
  geom_vline(aes(xintercept = mean(danceability) + sd(danceability)), color = "blue", linetype = "dashed", size = 1)
```

### Task 7: Describe our results and pitch a story

Write a few sentences about the distribution of danceability scores in the Spotify dataset. Is it a normal distribution, or is it skewed in one direction? What does the mean and one standard distribution tell us about the songs in the dataset? Draft a short story pitch from this data.

The danceability scores are skewed left with a mean score of around 0.67 and standard deviation of around 0.16. More of the data is clustered toward the center/mean, and around 68% of it falls within 1 standard deviation point. This also means that 34% of the Spotify songs are distributed above the mean and have a greater danceability value. 

I'm wondering what goes into measuring a song's "danceability," and what does "danceability" even mean? If a song has a danceability score closer to 1, does that mean it's more popular? Are Spotify users the people measuring a song's danceability, or is it random people from a survey? At the moment, this data doesn't mean much and there is no background information. However, if I were to pitch a story from this data, I would want to know what went into ranking these particular songs by "danceability," and whether Spotify users (potential interviewees) would agree with a song's specific danceability score.

A lot of the songs in the dataset have a greater danceability value. 
### Task 8: Save Your Work and Put it on GitHub

Save this notebook, then commit it to your local repository. Push the changes to your GitHub repository and submit the URL for this file in ELMS.
