---
title: "HW9_TestScores"
Name: Derek Willis/Rob Wells
Date: 10/21/2025
---

name: Lauren Frank 


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
```

## Did a New Reading Program Lead to Better Scores?

The superintendent recently claimed that a new reading program has improved third-grade reading scores across the school district.

Before the program, third-grade students in the district averaged 72.6 points on standardized reading tests with a standard deviation of 4.8 points.

After implementing the program for one semester, you collected scores from 12 randomly selected classrooms:
74, 76, 73, 75, 78, 77, 74, 79, 75, 76, 77, 75

As a journalist, you need to determine: **Is there statistical evidence that reading scores have actually improved?**

## Task 1: Organize your data and initial assessment

Before you can run this codeblock, you will need to fill in a value where it says REPLACE_ME. That value can be found in the introduction.

```{r}
# Known information about reading scores before the new program
prior_mean <- 72.6  # average score
prior_sd <- 4.8    # standard deviation

# Reading scores after implementing the new program (12 classrooms)
new_scores <- c(74, 76, 73, 75, 78, 77, 74, 79, 75, 76, 77, 75) # Replace with the actual scores

# Create a journalist-friendly dataset
score_data <- tibble(
  classroom = paste("Classroom", 1:12),
  reading_score = new_scores
)

# View the data
score_data
```

### Reflection Question 1:  (2 points)
Based on just looking at the score_data dataframe, have test scores improved? How can you tell?

Just looking at the score_data dataframe, I can tell that the test scores improved because all of the new reading scores are greater than 72.6 points, the previous average of standardized reading tests.


## Task 2: Calculate key statistics  (2 points)

Like Task 1, you will need to replace values where it says REPLACE_ME before running any code.


```{r}
# Calculate statistics based on the new reading scores
new_stats <- score_data |> 
  summarise(
    mean= mean(new_scores),
    sd = sd(new_scores),
    n = n()
  )
new_stats
```

### Reflection Question 2: (3 points)
Looking at the mean and standard deviation of the new scores compared to the previous statistics, what initial observations can you make? Provide examples from the data in your answer. What questions might these statistics raise for your reporting?

- The mean has increased. The mean of the new scores is 75.75. This value is greater than that of the previous mean which was 72.6. 
- The standard deviation has decreased. The standard deviation of the new scores is 1.8. This value is smaller than the previous standard deviation which was 4.8. Since the standard deviation is a smaller, the data points are more clustered towards the mean than previously. This indicates that the scores are more consistent with the average.  

Questions:
- How did the new reading program help increase the test scores? What specific elements of the reading program may have contributed to these improved scores? 

## Task 3: Create a column chart  (1 points)

As before, replace any values marked REPLACE_ME based the instructions.


```{r}
# STUDENT TASK: Choose an appropriate fill color for the bars
my_fill_color <- "royalblue" # Replace with a color name like "royalblue", "darkgreen", etc.

# Create a visualization comparing new scores to the previous average
score_data |> 
ggplot(aes(x = classroom, y = reading_score)) +
  geom_col(fill = my_fill_color, alpha = 0.8) +
  geom_hline(yintercept = prior_mean, color = "darkred", size = 1, linetype = "dashed") +
  annotate("text", x = 2, y = prior_mean - 1, 
           label = "Previous Average (72.6)",  hjust = -0.5, vjust= 2, fontface = "bold", color = "darkred") +
  labs(
    title = "Reading Scores After New Program Implementation",
    subtitle = "Horizontal line shows previous district average of 72.6 points",
    x = NULL,
    y = "Reading Test Score",
    caption = "Source: District Assessment Data"
  ) +
  theme_minimal() +
  theme(
    plot.title = element_text(face = "bold", size = 14),
    plot.subtitle = element_text(size = 10),
    axis.text.x = element_text(angle = 45, hjust = 1)
  )
```

### Reflection Question 3: (3 points)
Examine the chart you created, and suggest a better title based on the results of the data, not a description.
"Average Reading Scores Increase by 3 points in Classrooms After Implementation of New Reading Program."

## Task 4: Perform a hypothesis test (2 points)

This is where we formally test the superintendent's claim that reading scores have improved. Fill in the REPLACE_ME values as needed, beginning with your hypotheses.

**Hypotheses:**
Null: The new reading program has not increased reading scores
Alternative: The new reading program has increased reading scores

```{r}
# Set the significance level for your test
alpha_level <- 0.05 # Replace with the appropriate value

# Perform a one-sample t-test
# Since we want to know if scores improved (increased), we use a one-sided test (alternative = "greater")
t_test_result <- t.test(
  score_data$reading_score,
  mu = prior_mean,
  alternative = "greater"
)

# Display the results
t_test_result
```

### Reflection Question 4:  (3 points)
What does the p-value tell you, and what doesn't it tell you? How would you explain these results to a non-technical audience while maintaining accuracy?

The p-value (0.00003435) is less than the alpha level (0.05). This means that the alternative hypothesis is true, so the reading scores did improve after the implementation of the new program. 

## Task 5: Interpreting the results for your news story (2 points)

Let's gather all of the important stats we'll need in one place, so we can look at the prior average, the new scores and the results of the t.test, including the confidence interval. Replace any values where it says REPLACE_ME.


```{r}
# Get the p-value
p_value <- t_test_result$p.value

# Calculate the 95% confidence interval
ci <- t.test(score_data$reading_score)$conf.int

# Create a tibble to display the key statistics for your story
story_stats <- tibble(
  `Previous average` = prior_mean,
  `New average` = mean(75.75),
  `Improvement` = mean(new_scores) - prior_mean,
  `Percent change` = round(((mean(new_scores) - prior_mean) / prior_mean) * 100, 1),
  `p-value` = p_value,
  `Lower bound` = ci[1],
  `Upper bound` = ci[2],
  `Confidence level` = "95%"
)

# Display the key statistics
story_stats
```

## Conclusion

### Reflection Question 5: (4 points)
Based on these statistics, what would be your headline and lead paragraph for this story? Is there evidence to support the superintendent's claim?
- There is evidence to support the superintendent's claim.
- Headline: "3rd Grade Reading Scores Improve After Implementation of New School Program."
- Lead: A recent analysis of data from 12 randomly selected classrooms shows that 3rd grade reading scores improved following the implementation of the district’s new reading program. 

### Reflection Question 6:  (1 point)
What metrics or outcomes beyond test scores might be important to track for assessing reading performance?
- Depends on each student and how well they can read and comprehend information. 
- The teacher or teacher's teaching style and how well that helped the student. 
- What did the reading program look like, and did it vary across the classroom's in the school district? 
